One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-22 papers/1_dmatpol/paper-r2.tex
Revision in second iteration: Notation regularized and added subsection in the discussion

We introduce the necessary theory and derive the \emph{ab initio} density matrix susceptibility model
at the HF approximation\cite{Roothaan.RevModPhys.1951} 
in Section~\ref{s:2}. Subsequently, this model serves as a basis for generalization to the 
arbitrary one\hyp{}electron density with electronic correlation which is derived in Section~\ref{s:2}. 
The implementation details in a computer code as well as the validation procedures are detailed
in Section~\ref{s:3}. We then discuss the performance of the models in Section~\ref{s:4} 
%based on a few small molecules
%in varying electric field distributions. 
Finally, in Section~\ref{s:5} we conclude the presentation with a few 
remarks for future development.

%
%Understanding intricate mechanisms that govern
%the biomachinery of life requires thorough

@@

The electronic energy of a closed\hyp{}shell molecule is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho-1}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
The electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state. 
In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).

In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are probably not sufficient enough to accurately describe the local polarization
of electron density distribution when the electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed site model at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this Work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

@@

         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
  +\Tr{ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
  & \\ \nonumber =

@@

   \right] }  \;,
\\
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}

@@

 +\Tr{ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
  & \\ \nonumber =

@@

   \widetilde{\mathbb{M}}  &\equiv {\bf C}^\dagger {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^\dagger {\mathbb{M}} {\bf D}^{(0)} \;, \\

@@

      = \left[  {\bf C}^\dagger \mathbb{M} \left( \bf{1} - \bf{D}^{(0)} \right) \right]_{i\alpha}

@@

       \left[ {\bf L}_i^T {\bf L}_i \right]^{-1} {\bf L}_i^T 

@@

does not exist because ${\bf L}_i {\bf L}_i^T$ is singular due to the fact that usually

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parametereized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w} = \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i)}_{\alpha\beta} = \Big\{
                               C_{\alpha i} {\bf b}_{\beta i} + C_{\beta i} {\bf b}_{\alpha i} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}
%
\Big\}  \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i)}$
are unique properties of electronically unperturbed molecule.

One very important feature of the susceptibility from Eq.~\eqref{e:final-model.HF} is that
the density matrix increment is dependent \emph{only} on the susceptibility element associated with the same
pair of AO's. Therefore, there is no mixing in the AO's, rather the increment is proportional to the 
vector susceptibility, different for each density matrix element.

@@

The exact mathematical form of the generalized density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown. 
The susceptibility from Eq.~\eqref{e:final-model.HF} is valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarizability approximation), 
but not higher multipole moments, are correctly described.
Our strategy is to treat the susceptibilities from Eq.~\eqref{e:final-model.General}
as adjustable parameters, rather than try to derive them from first principles theory.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule, energy\hyp{}optimized by using the HF/6-311++G** 
method\cite{Roothaan.RevModPhys.1951,Krishnan.Binkley.Seeger.Pople.JCP.1980}, 
was selected as a test system which was subject to 
electrostatic perturbations due to randomly selected uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields. 
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval of $\vert 0.02 \vert$ a.u.
generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of perturbing charges were drawn from the uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).
In this way, independent sets of 100 random samples of uniform electric fields as well as 
40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.

>>>>

We introduce the necessary theory and derive the \emph{ab initio} density matrix susceptibility model
at the HF approximation\cite{Roothaan.RevModPhys.1951} 
in Section~\ref{s:2}. Subsequently, this model serves as a basis for generalization to the 
arbitrary one\hyp{}electron density with electronic correlation which is derived in Section~\ref{s:2}. 
The implementation details in a computer code as well as the validation procedures are detailed
in Section~\ref{s:3}. We then discuss the performance of the models in Section~\ref{s:4}.
Finally, in Section~\ref{s:5} we conclude the presentation with a few 
remarks for future development.

@@

The electronic energy of a closed\hyp{}shell molecule is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho-1}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
The electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state.

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are probably not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this Work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory,\cite{Roothaan.RevModPhys.1951} ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C} \cdot {\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right] \cdot
                                   \left[ {\bf 1} + {\bf v}^\dagger \cdot {\bf v} \right]^{-1} \cdot
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] \cdot {\BM\Delta} \cdot {\bf D}^{(0)} + 
                        {\bf D}^{(0)} \cdot {\BM\Delta}^\dagger \cdot \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C} \cdot {\bf C}^{(0)\dagger} + {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \nonumber \\
           &\qquad - {\bf D}^{(0)} \cdot \delta {\bf C} \cdot {\bf C}^{{(0)}\dagger} 
                   - {\bf C}^{(0)} \cdot \delta {\bf C}^\dagger \cdot {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger} \cdot {\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated otherwise.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This guarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

%\begin{subequations}
% \begin{align}
%  \Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}  
%    \right] }  &
%  \nonumber\\ 
%  +\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger}
%    \right] }
% &=
%2 \Tr{ 
%    \left[ 
%         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
%   \right] }  \;, 
%\\
%  \Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta} \cdot {\bf D}^{(0)}
%    \right] }  &
%   \nonumber\\
% +\Tr{ 
%    \left[ 
%         {\mathbb{M}} \cdot {\bf D}^{(0)} \cdot {\BM\Delta}^{\dagger} \cdot {\bf D}^{(0)}
%    \right] }
% &= 
%2 \Tr{ 
%    \left[ 
%         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
%   \right] } \;.
% \end{align}
%\end{subequations}

@@

 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong

@@

@@

-

@@

   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\

@@

      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}

@@

       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T 

@@

does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i}^{(10)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w}^{(10)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(10)$' denotes the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field and zeroth\hyp{}order with respect to
the electric field gradient.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;10)}_{\alpha\beta} = \Big\{
                               C_{\alpha i} {\bf b}_{\beta i}^{(10)} + C_{\beta i} {\bf b}_{\alpha i}^{(10)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}^{(10)}
%
\Big\}  \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;10)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;10)}$
are unique properties of electronically unperturbed molecule.

One very important feature of the susceptibility from Eq.~\eqref{e:final-model.HF} is that
the density matrix increment is dependent \emph{only} on the susceptibility element associated with the same
pair of AO's. Therefore, there is no mixing in the AO's, rather the increment is proportional to the 
vector susceptibility, different for each density matrix element. This feature will be used to
generalize the model to an arbitrary level of theory and electric field curvature in the next section.

@@

The exact mathematical form of the generalized density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown. 
The susceptibility from Eq.~\eqref{e:final-model.HF} is valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarizability approximation), 
but not higher multipole moments and polarization energy, are correctly described.
Our strategy is to treat the susceptibilities from Eq.~\eqref{e:final-model.General}
as adjustable parameters, rather than try to derive them from first principles theory.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule
%, energy\hyp{}optimized by using the HF/6-311++G** 
%method\cite{Roothaan.RevModPhys.1951,Krishnan.Binkley.Seeger.Pople.JCP.1980}, 
was selected as a test system which was subject to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields. 
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval of $\vert 0.02 \vert$ a.u.
generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of perturbing charges were drawn from the uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).
In this way, independent sets of 100 random samples of uniform electric fields as well as 
40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.
